Making Sequential Programs Deterministic

### 2019-03-19 introduction.tex
Edit first half of introduction

Writing correct programs is hard. Programmers often write correctness specifications for their code in the form of assertions. They write test cases that cover several usage scenarios to strengthen their belief that
the code works as intended. Despite all the testing effort, it often happens that a test suite produces
diverging results for different users or for the same user when the test suite is run again. When a test suite fails
nondeterministically, it  reduces the programmers' confidence in the program. Additionally, reproducing
such bugs is difficult making debugging an onerous task. Examples of the underlying causes of such flakiness 
could be OS scheduling in case of concurrent programs, use of an API that has nondeterministic specification,
accessing system properties such as the file system or environment variables. In this paper, we focus on
nondeterminism arising from the use of APIs that have nondeterministic specifications in java programs.
Instances of such APIs in the JDK include Date(), Random(), and HashSet iteration. Note that mere calls
to these APIs is not problematic. But if the programmer intends the results of these calls to be deterministic,
it could result in undesirable results. We aim to capture this behavior and report it. The high level goal
of our work is to provide programmers with a tool for specifying  deterministic properties in a program
and be able to verify them at compile time.

Recent work in this area~\cite{nondex,deflaker} has focussed its efforts on identifying flaky test cases.
NONDEX~\cite{nondex} is a tool that manually identifies methods in the JDK that have nondeterministic specifications, builds models for
these methods that behave nondeterministically, and  use sa modified JVM that runs the test suite multiple times.
Diverging test results in this setting indicate a flaky test. DeFlaker~\cite{deflaker} looks at a range of commit versions
of a code, and marks a test as flaky if it doesn't execute any modified code but still fails in the newer version. These techniques
have been able to identify issues in real world programs, some of which have been fixed by the developers. This shows that nondeterminism
is an important problem and solving it at an earlier phase in the software development lifecycle would be greatly beneficial to 
developers.

@@

Our type system hierarchy consists of three types: a) @Det which represents values that will be the same across executions,
b) @OrderNonDet which represents collections that are guaranteed to contain the same elements albeit in possibly
different iteration order, and c) @NonDet which provides no determism guarantees.  We found this type hierarchy 
is expressive enough to identify nondeterministic behaviors in production software. We annotated all the collection
classes (Lists, Sets and Maps) in the JDK with our new type system. Our tool is therefore guaranteed to report any violation
of deterministic properties in a client program caused by methods in these classes.

We make the folowing contributions through this paper:
\begin{enumerate}
	\item We provide a design and implemention of a type system for expressing deterministic properties in sequential programs.
	\item We annotated several classes in JDK. By doing this, we have translated the specification for these APIs from
	human readable comments to machine readable and verifyable documentation.
	\item We performed a case study of a real world java program Randoop (loc, etc) which revealed bugs.
	\item We perform a comparative study of our technique with state of the art flaky test detectors and present results.
\end{enumerate}

>>>>

A nondeterministic program may produce different output on different runs
when provided with the same input.
Nondeterminism is a serious problem for software developers.
\todo{Add lots of citations in the below.}
\begin{itemize}
\item
  Nondeterminism makes a program difficult to \textbf{test}, because test
  oracles must account for all possible behaviors while still enforcing
  correct behaviors.  Test oracles that are too strict lead to flaky tests
  that sometimes pass and sometimes fail.  Flaky tests must be re-run, or
  developers ignore them; in either case, their utility to detect defects
  is limited.
\item
  Nondeterminism makes it difficult to \textbf{compare} two runs of a
  program on different data, or to compare a run of a slightly modified
  program to an original program.  This hinders debugging and maintenance,
  and prevents use of techniques such as Delta Debugging~\cite{Zeller1999}.
\item
  Nondeterminism reduces users' and developers' \textbf{trust} in a program's output.
\end{itemize}

Two well-known sources of nondeterminism are concurrency
% (due to OS scheduling or message delivery over a network)
and coin-flipping
(calls to a \<random> API).
It may be surprising that nondeterminism is common even in sequential
programs that do not flip coins.
For example, any program that iterates over a set or map, and that uses the
results to affect the output, may suffer from nondeterminism.
Other nondeterministic APIs include date-and-time functions and
accessing system properties such as the file system or environment variables.

We have created an analysis that detects nondeterminism so that
programmers can eliminate it.
The analysis permits programmers to specify which parts of their program
are nondeterministic, and it verifies that the remainder is deterministic.
%
The analysis is sound:  if it issues no warnings, then the program does not
suffer nondeterminism of the given types.  (If the program uses concurrent,
it might still be nondeterministic.)
%
The analysis works at compile time, giving a guarantee over every possible
execution of the program, by contrast to unsound dynamic tools that attempt
to discover when a program has exhibited nondeterministic behavior on
specific runs.
%
The analysis differentiates between values that can differ arbitrarily on
different runs, and collections that will contain the same values, but
possibly in a different order.
%
The analysis is precise enough for practical use.  It permits calls to
nondeterministic APIs, and only issues a warning if they are used in ways
that may lead to nondeterministic output observed by a users.  Like any
sound analysis, it can issue false positive warnings.

The high level goal of our work is to provide programmers with a tool for
specifying deterministic properties in a program and be able to verify them
at compile time.

Previous work in program analysis for nondeterminism has focused on unsound dynamic
approaches that identify flaky test cases.
NonDex~\cite{nondex} is a tool that manually identifies methods in the JDK that have nondeterministic specifications, builds models for
these methods that behave nondeterministically, and  use a modified JVM that runs the test suite multiple times.
Diverging test results in this setting indicate a flaky test. DeFlaker~\cite{deflaker} looks at a range of commit versions
of a code, and marks a test as flaky if it doesn't execute any modified code but still fails in the newer version. These techniques
have been able to identify issues in real world programs, some of which have been fixed by the developers. This shows that nondeterminism
is an important problem and solving it at an earlier phase in the software development lifecycle would be greatly beneficial to 
developers.

@@

Our type system hierarchy consists of three types: a) @Det which represents values that will be the same across executions,
b) @OrderNonDet which represents collections that are guaranteed to contain the same elements albeit in possibly
different iteration order, and c) @NonDet which provides no determinism guarantees.  We found this type hierarchy 
is expressive enough to identify nondeterministic behaviors in production software. We annotated all the collection
classes (Lists, Sets and Maps) in the JDK with our new type system. Our tool is therefore guaranteed to report any violation
of deterministic properties in a client program caused by methods in these classes.

We make the following contributions through this paper:
\begin{enumerate}
	\item We provide a design and implementation of a type system for expressing deterministic properties in sequential programs.
	\item We annotated several classes in JDK. By doing this, we have translated the specification for these APIs from
	human readable comments to machine readable and verifiable documentation.
	\item We performed a case study of a real world java program Randoop (loc, etc) which revealed bugs.
	\item We perform a comparative study of our technique with state of the art flaky test detectors and present results.
\end{enumerate}

% LocalWords:  NONDEX NonDex DeFlaker
